# Longitudinal Data Anlaysis 

Wide data has a column for each variable whereas long format data has a column for possible variable type & a column for the values of those variables


```{r}
# libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(patchwork)
library(rstatix)
library(ggpubr)
```


## Chapter 8: Analysis of Longitudinal Data I: Graphical Displays Summary Measure Approach

Data from a nutrition study conducted in three groups of rats wll be used. the groups were put on different diets and each animals body weight (grams) was recorded repeatedly (approximately) weekly, expect in week seven when two recording were taken) over a 9-week period. 

#### Load the data and check the variables and structure
```{r}
#set working directory
setwd("C:/LocalData/ocholla/IODS-project/Data")

RATS<- read.table("Rats.txt", header = TRUE, sep = '\t')

#convert variables ID and Group to factors
RATS<- within(RATS, {
  ID<- factor(ID)
  Group<- factor(Group)
})

#Convert data from wide to long form
RATSL<- RATS%>% 
  gather(key = WD, value = Weight, -ID, -Group) %>% 
  #mutate a new variable Time by extracting the number of the day from WD
  mutate(Time = as.integer(substr(WD,3,4)))
glimpse(RATSL)

```


#### Plot the RATSL values for all 11 rats mean, differentiating between the  nutrition groups into which the rats have been randomized

```{r}
ggplot(RATSL, aes(x=Time, y= Weight, linetype = WD))+
  geom_line()+
  scale_linetype_manual(values = rep(1:10,
                                     times=4))+
  facet_grid(.~ Group, labeller = label_both)+
  theme(legend.position = "none")+
  scale_y_continuous(limits = c(min(RATSL$Weight),
                                max(RATSL$Weight)))

```

From the graph, the rats weights scores varied within the groups across the 9 weeks. The rats with higher weight scores at the begining tend tp have higher scores throughout the study.   

* In group 1, the rats had the lowest weight across the period compared to group 2 and 3.  
* In group 2: the weights variation was the largest across the period compared to the other groups.    
* In group 1: the weight was relatively constant across the period


#### Standardize values of each observation (__tracking phenomenon__)

Observing the weights of rats from the beginning to the end of the study are clearly displayed through tracking phenomenon.  This tracking s done by standardizing the weight scores across the study period.  Plot the standardized observations aganist the study period

```{r}
#standardize the variables weights
RATSL<- RATSL %>%
  group_by(Time) %>% 
  mutate(stdWeight = (Weight - mean(Weight))/ sd(Weight) ) %>% 
  ungroup()

#plot again with the standardized weights

ggplot(RATSL, aes(x = Time, y = stdWeight,
                  linetype = WD))+
  geom_line()+
  scale_linetype_manual(values = rep(1:10,
                                     times = 4))+
  facet_grid(. ~Group, labeller = label_both)+ scale_y_continuous(name = "standardized Weights")

```

From the plot:  

* The mean of group 2 and group 3 at 1, suggesting a small difference between the two groups.  
* The mean of group 1 is at -1, this diverges from the rest, indicating the difference between group 1 and the other two groups.


With many observations, graphical display of individual observation are often obscured making them little of importance. This creates a need to produce graphs showing average profiles for each nutrition group along with some indication of the variation of the observation at each week point

```{r}
#number of subjects (days), baseline (week 0) included
n<- 16

#summary data with mean and standard error of RATS by nutrition group and Time
RATSS<- RATSL %>%
  group_by(Group, Time) %>% 
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()
#plot the mean profiles
ggplot(RATSS, aes(x= Time, y= mean, linetype = Group, shape = Group))+
  geom_line()+
  scale_linetype_manual(values = c(1,2,3))+
  geom_point(size =3)+ #change from 3 to 2
  scale_shape_manual(values = c(1,2,3))+
  geom_errorbar(aes(ymin= mean -se, ymax= mean + se,
                    linetype ="1"), width= 0.3)+
  theme(legend.position = "right")+
  scale_y_continuous(name = "mean(Weight)+/- se(Weight)")

```

There is no overlap in the mean profiles of the three nutrition groups indicating the difference between the three groups with respect to the mean of rats weight.  


Possible alternative is plotting the mean profiles of the three groups to a side by side box plots of the weights at each week weight score.

```{r}
ggplot(RATSL, aes(x= factor(Time), y = Weight, 
                  fill = Group))+
  geom_boxplot(position = position_dodge(width = 0.9))+
  theme_bw()+theme(panel.grid.major = element_blank(), 
                   panel.grid.minor = element_blank())+
  theme(legend.position = "bottom")+ #c(0.8,0.8)
  scale_x_discrete(name = "Time")

```

The plot suggest possibility of outliers in the weights, also the plot indicates again the general increase of rats weight across the study period in all the three groups

#### Applying the summary measure approach

The mean of day 8-64 will be the chosen summary measure. This measure is calculated and then displayed using boxplots for each nutrition group.


```{r}

RATSL8S<- RATSL %>%
  filter(Time > 1 ) %>% # 1 is the baseline date
  group_by(Group, WD) %>%
  summarise( mean = mean(Weight) )%>%
  ungroup()
#draw  a box plot of the mean versus Group
ggplot(RATSL8S, aes(x = Group, y =mean)) +
  geom_boxplot()+
  stat_summary(fun.y = "mean", geom = "point", shape = 23, size =4, 
               fill = "blue")+
  scale_y_continuous(name = "mean(Weight), Time 8-64")

```


From the boxplot:  

* The mean summary measure is not varying across the three diet groups, there seems to be no skew observations.  
* There are no presences of outliers across the three diet groups.  

With no outliers the subsequent tasks on removing outliers are not applicable

#### Apply *t*-test to assess difference between the nutrition groups


```{r}
res.aov<- RATSL8S %>% anova_test(mean ~ Group)
res.aov

```

From the results, there are significant difference between the three deit groups as pvalue <0.05 (P= 1.5e-26) which is highlighted by "*", F(2,27)=1091, eta2[g]=0.988.  

#### using baseline data

The use of baseline data helps in gaining precision when it is used appropriately as a co variate in an analysis of covariance 



```{r}


```


## Chapter 9: Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables

Longitudinal data, where a response variable is measured on each subject on different occasions poses a challenge as repeated measurement often are likely correlated rather than independent. 

To test how linear mixed effect models are applied data taken from Davis 2002 will be used. It comprises of 40 male subjects who were randomly assigned to **one or two treatment** groups and each subject was rated on the **Brief Psychiatric Rating Scale (BPRS)** measured before treatment began **(week 0)** and then at weekly interval for **eight weeks**. The BPRS assesses the level of 18 symptoms construct such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not presented) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.
The question of most interest is whether BPRS differs in the two treatments differ
 
### 9.3.1 Fitting the independence Model to the BPRS Data

Load the data, ignoring the repeated-measure structure of the data and assume that all the observation are independent of one another 
```{r}
#set working directory
setwd("C:/LocalData/ocholla/IODS-project/Data")
#Load the data sets (BPRS)
BPRS<- read.table("BPRS.txt", sep= " ", header = TRUE)
#convert subject and treatment from integer to factors
BPRS<- within(BPRS, {
  treatment<- factor(treatment)
  subject<- factor(subject)
})

#Add column on the number of count of men 
BPRS$ID<- seq.int(nrow(BPRS))
glimpse(BPRS)

```

The data is in wide form format, it needs to be converted to long form to be used for analysis


1. Convert the data to long form
```{r}

BPRSL<- gather(BPRS, key = weeks, value = bprs, week0:week8) %>%
  #extract the week number- adding the variable week into the dataset
  mutate(week = as.integer(substr(weeks, 5, 5)))
#take a glimpse at the BPRSL data
glimpse(BPRSL)

```
When we ignore the set of 9 bprs measurements come from the same subject, we have 360 bprs, weeks and treatment that can be easily analysed by using multiple linear regression


Plot the data, identifying the bprs observation in each treatment but ignoring the longitudinal nature of data
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = ID))+
  geom_text(aes(label = treatment))+
  #geom_line(aes(linetype = subject))+
  scale_x_continuous(name = "Week (number)", breaks = seq(0,8,1))+
  scale_y_continuous(name = "bprs ()") + 
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = 
          element_blank())

```

In general, BPRS decreases in both treatment 1 &2  as the number of weeks increased. 
Men who have higher BPRS values at the beginning tend to have higher values throughout the study.


#### Fit a Linear regression model 

Fit a multiple linear regression model with bprs as response variable while week and  treatment as explanatory variables, and ignpring the repeated-measures structure of the Data

```{r}

#create a regression model BPRS_reg
BPRS_reg<- lm(bprs ~ week + treatment, data = BPRSL)
#summary 
summary(BPRS_reg)
```

From the model, the week variable is highly significant (p<0.5) and negatively correlated to BPRS measure. Implying that with increase in weeks, the BPRS decreased.
Treatment 2 is not significant when holding Treatment 1 as a constant.
The two explanatory variables explained approximately 18% of BPRS variability.
This model (multiple linear regression) assumes independence of the repeated measures of BPRS, and thus assumption is highly unlikely.

### 9.3.2 Fitting Linear Mixed Models to the BPRS data

A graphical display of the bprs data taking into account the longitudinal structure of the data by joining together the bprs measures belonging to each man. 

```{r}

ggplot(BPRSL, aes(x= week, y = bprs, group= ID))+ 
  geom_line(aes(linetype= treatment))+
  scale_x_continuous(name= "Weeks (number)", breaks = seq(0,8,1))+
  scale_y_continuous(name = "bprs")+
  theme_bw()+ theme(legend.position = "top")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = 
          element_blank())

```

From the plot, there is a general decline in bprs across all men as the number of weeks increased across both treatments.


A scatterplot matrix of repeated measures of BPRS to assess the independence of the repeated measures

```{r}
pairs(BPRS[, 3:11], cex = 0.7)

```

From the scatter plot: The repeated measures of bprs are not independent of one another.They are correlated with one another.

#### Fitting Random intercept Model , with week and treatment as Expanatory variables

First fir the random intercept model and include the two explanatory variables- week and treatment. This kind of model allows the linear regression fir for each man to differe in intercept from other men

```{r}
#create a random intercept model
BPRS_ref<- lmer(bprs ~ week + treatment + (1 | ID), data =  BPRSL, REML = FALSE)

#print the summary of the model
summary(BPRS_ref)

```

The estimated variance of the subject random effects (97.39) is not very large indicating a small variation in the intercept of the regression fits of each subject (men) treatment profile. The estimated standard error of **week** is smaller when fitting random intercept compared to linear regression model.This suggest that increase in weeks leads to ignoring within-subject dependences, which reduces the error variance in the model.

#### Fit the Random intercept and slope model , with week and treatment as explanatory variables

The effect from the random intercept in this model allows the linear regression fits for each individual to differ in slope

```{r}
BPRS_ref1<- lmer(bprs ~ week + treatment + (week|ID), data = BPRSL, REML = FALSE)
#print a summary of the model
summary(BPRS_ref1)

#perform an ANOVA test on the two models
anova(BPRS_ref1,BPRS_ref)

```


The above model  gives a lower likelihood ratio test, AIC and BIC values compared to the earlier Random intercept model. Furthermore is has a small P value which is significant, these results indicates that random intercept and slope model provides a better fit for the data.



#### Fitting a random intercept and slope model that allows for a treatment x week interaction


```{r}

BPRS_ref2<- lmer(bprs ~ week * treatment + (week|ID),
                 data = BPRSL, REML = FALSE)

#print a summary of the model
summary(BPRS_ref2)

#perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref2)

```

The likelihood ratio test of the interaction random intercept and slope model against the corresponding model without interaction is -1253.7 with 2 degree of freedom (DF), with a big P-value that is not statistically significant. From these results, the interaction model  does not provides a better fit for the men treatment data.


#### Find the fitted values from the interaction model and plot the fitted growth rates for each rat


```{r}
#create a vector of the fitted values
Fitted<- fitted(BPRS_ref2)
#Create a new column fitted to BPRSL
BPRSL<- BPRSL%>% 
  mutate(Fitted)

#plot for the observed values
p1<- ggplot(BPRSL, aes(x = week, y = bprs, group = ID))+  
  geom_line(aes(linetype = treatment))+ 
  scale_x_continuous(name = "Week (number)", breaks = seq(0, 8,1))+
  scale_y_continuous(name = "Observed bprs ")+
  theme_bw()+ theme(legend.position = "right")+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
  ggtitle("observed")

#plot for the fitted values
p2<-ggplot(BPRSL, aes(x = week, y= Fitted, group = ID))+
             geom_line(aes(linetype = treatment))+
             scale_x_continuous(name = "Week (number)", breaks = seq(0,8,1))+
             scale_y_continuous(name = "Fitted bprs ")+
             theme_bw()+theme(legend.position = "right")+
             theme(panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank())+
             ggtitle("Fitted")

#plot the observed and fitted
p1 +p2
```

The plot underlines how the interaction models predicted the observed data. The fitted plot includes the random effects of the model


```{r}

date()

```
