# Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
#set the working directory of your R session the IODS project folder
setwd("Z:/Courses/Open data science/R/IODS-project")

students2014<-read.csv("Data/learning2014.csv", header = TRUE, stringsAsFactors = TRUE)
```

Explore the data structure and its dimensions

```{r}
str(students2014)
dim(students2014)


```
The data contains 166 observations of 7 variables. 
Two of the factors are integers (age and points), four numericals (attitude, deep, stra and surf), while gender variable is a factor

```{r}



```


### Graphical overview and summaries of the variables in the data
```{r}
summary(students2014)

```
The summary shows the descriptive statistics of the numerics and integers, while showing the distribution of male and female in the gender variable

#### visualizing  student attitude aganist exam points
```{r}
library(ggplot2)
#initialize plot with data and aesthetic mapping
p1<- ggplot(students2014, aes( x= attitude, y= points, col = gender))
#define the visualization type (points)
p2<- p1 + geom_point()
#draw the plot
p2
#add a regression line
p3<- p2+geom_smooth(method = "lm")
#add a main title and draw the plot
p4<- p3 + ggtitle("student's attitude versus exam points")
p4
```

### Scatterplot matrix of the variables in the data
```{r}
# exclude the gender variable from the matrix
pairs(students2014[-1], col = students2014$gender)
#access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)
#create a more advanced plot matrix with ggpairs()
p<- ggpairs(students2014, mapping = aes (col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
#draw the plot
p

```



### fit a regression model
A regression model is divided into three parts and it follows the equation below
$$Y \sim X\beta_0+X\beta_1 +\epsilon $$  


  * where **Y** is the target variable 
  * $X\beta_0+X\beta_1$ is the systematic part of the model 
  * $\beta$ quantifies the realtionship between **y* and **x**.
  * $\epsilon$ describes the errors
  
The exaam point has been used as the target variables while surface learning, strategic learning and age have be used as dependent variables in the mode

```{r}
my_model<- lm(points~ surf+stra+age, data = students2014)

```
summary of the model
```{r}
summary(my_model)

```
comment and interpret the results.
Explain and interpret the statistical test related to the model parameters

using a summary of the fitted model, explain the relationship between the chosen explanatory variables and the target variable(interpret the model parameters). Explain and interpret the multiple R squared of the model

###Diagnostic plots
####Residual vs Fitted values
```{r}
summary(my_model)

```
Explain the assumption of the model and interpret the validity of those assumptions based on the diagnostic plots

####Normal QQ-plot and Residuals vs Leverage
```{r}
summary(my_model)

```
Explain the assumption of the model and interpret the validity of those assumptions based on the diagnostic plots



Here were fo
